Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-28 paper.qmd
Update paper.qmd

A more robust approach than simply listing package versions is to use tools that
define per-project environments. This is common for [Python]{.proglang} users,
who have many tools at their disposal, such as the built-in *virtual
environments* module, or third-party tools like [uv]{.pkg}, which manages both
project-specific interpreters and package libraries using a lock file mechanism.

@@

[Nix]{.pkg} is a cross-platform package manager designed for reproducible
software installation and deployment. Unlike traditional package managers,
[Nix]{.pkg} emphasizes immutable infrastructure and a functional programming
paradigm to ensure computational environment consistency. The primary
[Nix]{.pkg} package collection, `nixpkgs`, provides access to over 120,000
packages, including nearly all of CRAN and Bioconductor.^[Only about 80 packages 
currently fail to build on the NixOS build farm, plus roughly 30 packages marked as broken
by the maintainers of the [R]{.proglang} ecosystem. These could, in principle, be fixed, 
but doing so would require significant effort. Marking them as broken signals to users 
that their build failures are known and intentional.] According to Repology,
a service tracking package repositories across distributions, `nixpkgs` is both
the largest and among the most up-to-date repositories available (Figure
@fig-repology).

@@

A drawback of forking `nixpkgs` is that backported packages are not included
upstream and thus are not prebuilt by Nix’s CI platform, Hydra. Users may need
to build many packages from source, which can be time-consuming. To mitigate
this, we provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
complementing the public Nix cache. Instructions for using Cachix are in
[rix]{.pkg}’s documentation. Using the cache significantly speeds up
installations, as prebuilt packages are downloaded rather than compiled.

@@

To illustrate the practical benefits of our framework, we implemented the RBC
pipeline example from Section @sec-rixpress using two distinct paradigms. The
first is a traditional, imperative, stack combining Docker for environment
isolation, Make for workflow orchestration, and language-specific package
managers. The second is our proposed declarative framework using `rix` and
`rixpress`. While both approaches successfully execute the same polyglot
analysis to produce identical outputs, their comparison illuminates fundamental
differences in conceptual models, system complexity, and the nature of the
reproducibility guarantees they provide.

@@

This declarative rigor, however, is not without pragmatic costs. Our validation
on GitHub Actions shows that the imperative Docker-based runner completes the
entire pipeline in approximately three minutes, whereas the declarative
Nix-based runner takes around five minutes. Nix’s deterministic dependency
resolution guarantees bit-for-bit reproducibility at the expense of initial
computation time, whereas the imperative system prioritizes immediacy over
long-term stability.

>>>>

A more robust approach than simply listing package versions is to use tools that
define per-project environments. This is common for [Python]{.proglang} users,
who have many tools at their disposal, such as the built-in *virtual
environments* module, or third-party tools like [uv]{.pkg} [@uv], which manages both
project-specific interpreters and package libraries using a lock file mechanism.

@@

[Nix]{.pkg} is a cross-platform package manager designed for reproducible
software installation and deployment. Unlike traditional package managers,
[Nix]{.pkg} emphasizes immutable infrastructure and a functional programming
paradigm to ensure computational environment consistency. The primary
[Nix]{.pkg} package collection, `nixpkgs`, provides access to over 120,000
packages, including nearly all of CRAN and Bioconductor.^[Only about 80 packages 
currently fail to build on the NixOS build farm, plus roughly 30 packages marked as broken
by the maintainers of the [R]{.proglang} ecosystem. These could, in principle, be fixed, 
but doing so would require significant effort. Marking them as broken signals to users 
that their build failures are known and intentional. It should be further noted that some
packages may build successfully, but then fail at runtime as they try to further install
runtime dependencies on first run. One such example is [torch]{.pkg}.] According to Repology,
a service tracking package repositories across distributions, `nixpkgs` is both
the largest and among the most up-to-date repositories available (Figure
@fig-repology).

@@

A drawback of forking `nixpkgs` is that backported packages are not included
upstream and thus are not prebuilt by Nix’s CI platform, Hydra. Users may need
to build many packages from source, which can be time-consuming. To mitigate
this, we provide a binary cache sponsored by 
[Cachix](https://www.cachix.org/)^[[https://www.cachix.org/](https://www.cachix.org/)],
complementing the public Nix cache. Instructions for using Cachix are in
[rix]{.pkg}’s documentation. Using the cache significantly speeds up
installations, as prebuilt packages are downloaded rather than compiled.

@@

To illustrate the practical benefits of our framework, we implemented the RBC
pipeline example from Section @sec-rixpress using two distinct paradigms. The
first is a traditional, imperative, stack combining Docker for environment
isolation, Make for workflow orchestration, and language-specific package
managers.^[This choice of tools is illustrative: we could have used, for instance,
[Snakemake]{.pkg} [@moelder2021snakemake] for workflow orchestration or [conda]{.pkg} [@anaconda] 
instead of [uv]{.pkg}, 
but the broader point remains—these are all imperative approaches] 
The second is our proposed declarative framework using `rix` and
`rixpress`. While both approaches successfully execute the same polyglot
analysis to produce identical outputs, their comparison illuminates fundamental
differences in conceptual models, system complexity, and the nature of the
reproducibility guarantees they provide.

@@

This declarative rigor, however, is not without pragmatic costs. Our validation
on GitHub Actions shows that the imperative Docker-based runner completes the
entire pipeline in approximately three minutes, whereas the declarative
Nix-based runner takes around five minutes.^[To run the benchmark yourself,
fork the paper’s repository, enable GitHub Actions, and push a single change to
trigger the runs.] Nix’s deterministic dependency
resolution guarantees bit-for-bit reproducibility at the expense of initial
computation time, whereas the imperative system prioritizes immediacy over
long-term stability.
